---
title: "Chapter 1: Your First Shiny App"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Create app directory and file
The simplest way to create a shiny app is by creating a new directory for your app and have a file called `app.R`. The `app.R` file will have 4 key components:

- `library(shiny)`
- the UI
- the server
- `shinyApp(ui, server)`

## Adding UI controls
Example UI
```{r}
ui <- fluidPage( # a layout function that sets up the basic visual structure of the app
  
  selectInput("dataset", label = "Dataset", choices = ls("package:datasets")), # an input control that lets the user interact with the app
  
  verbatimTextOutput("summary"), # an output control telling Shiny where to render the output
  
  tableOutput("table") # also an output control
)
```
The functions above all generate HTML. If you call them outside a shiny app, you'll get HTML printed out in the console


## Adding behaviour
We bring the outputs in the UI to life by defining them in the server. We tell shiny how to perform a computation rather than to do it.

Example server
We provide shiny with a recipe to produce outputs. We give those outputs an ID and place them in a list-like object called output. The recipe is wrapped in a render function, which produces a particular type of output. Render functions are paired with output functions from the UI.
```{r}
server <- function(input, output, session) {
  output$summary <- renderPrint({
    dataset <- get(input$dataset, "package:datasets")
    summary(dataset)
  })
  
  output$table <- renderTable({
    dataset <- get(input$dataset, "package:datasets")
    dataset
  })
}
```
As we tell the server how to produce outputs, they are reactive. As the values of inputs in the UI change, the outputs are recomputed.


## Reducing duplication with reactive expressions
The above server example contains duplication:
`dataset <- get(input$dataset, "package:datasets")`

In regular R scripting, we use either capture a value in a variable or capture the computation in a function. Neither of these approaches fully work with shiny so we use reactive expressions.

Reactive expressions are blocks of code wrapped in `reactive({...})` and assigning it to a variable. This variable is then called like a function.

Reactive expressions are called the first time it runs and then only when it needs to be updated because an input has changed.
```{r}
server <- function(input, output, session) {
  # Create a reactive expression
  dataset <- reactive({
    get(input$dataset, "package:datasets")
  })

  output$summary <- renderPrint({
    # Use a reactive expression by calling it like a function
    summary(dataset())
  })
  
  output$table <- renderTable({
    dataset()
  })
}
```

