---
title: "Chapter 18: Functions"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Writing functions in the UI and server have different flavours:
* In the UI, where there are components that are repeated in places with minor differences, pulling out the repeated code into a function reduces duplication and can be combined with functional programming techniques to generate many controls at once
* In the server, complex reactives are hard to debug. Pulling out reactives into separate functions makes it substantially easier to debug as you can experiment with comuptation independent of reactivity

Functions aren't just useful for reducing duplication and debugging purposes, they also allow you to spread out code across multiple files. 


## File Organisation
Recommended directory/file structure: 
* placing large functions in their own file labelled with its {name}.R
* placing smaller, simpler functions into one file (utils.R)


## UI Functions
Powerful tool to reduce duplication in UI code and make managing elements simpler and quicker.

Example: your UI needs a bunch of sliders that each range from 0-1, start at 0.5 and step 0.1. It could look like this:
```{r}
ui <- fluidRow(
  sliderInput("alpha", "alpha", min = 0, max = 1, value = 0.5, step = 0.1),
  sliderInput("beta",  "beta",  min = 0, max = 1, value = 0.5, step = 0.1),
  sliderInput("gamma", "gamma", min = 0, max = 1, value = 0.5, step = 0.1),
  sliderInput("delta", "delta", min = 0, max = 1, value = 0.5, step = 0.1)
)
```
But, as there are patterns here, we can create a function:
```{r}
sliderInput01 <- function(id) {
  sliderInput(id, label = id, min = 0, max = 1, value = 0.5, step = 0.1)
}

ui <- fluidRow(
  sliderInput01("alpha"),
  sliderInput01("beta"),
  sliderInput01("gamma"),
  sliderInput01("delta")
)
```
Now if changes need to be made, they just need to be made in one place, rather than over and over again.


### Other applications
Some examples of where functions could be creative ways of manipulating the UI.

A customised dateInput for a country:
```{r}
usWeekDateInput <- function(inputId, ...) {
  dateInput(inputId, ..., format = "dd M, yy", daysofweekdisabled = c(0, 6))
}
```

Radio buttons that make it easier to provide icons:
```{r}
iconRadioButtons <- function(inputId, label, choices, selected = NULL) {
  names <- lapply(choices, icon)
  values <- if (is.null(names(choices))) names(choices) else choices
  radioButtons(inputId,
    label = label,
    choiceNames = names, choiceValues = values, selected = selected
  )
}
```

### Functional programming
The slider input example could be further reduced.
```{r}
library(purrr)

vars <- c("alpha", "beta", "gamma", "delta")
sliders <- map(vars, sliderInput01)
ui <- fluidRow(sliders)
```
In this way, we use `map()` to call `sliderInput01()` once for each of the strings stored in var. sliders, which is a list, is then passed to `fluidRow()`, which unpacks the list so that the elements become children of the container.

[`map()` is equivalent to `lapply()`]

### UI as data
If the sliders we wanted had other differences, not just the id and label, we could create an inline data frame that defines the parameters of each control. 
```{r}
vars <- tibble::tribble(
  ~ id,   ~ min, ~ max,
  "alpha",     0,     1,
  "beta",      0,    10,
  "gamma",    -1,     1,
  "delta",     0,     1,
)
```
In the function we define, every argument name matches the column names
```{r}
mySliderInput <- function(id, label = id, min = 0, max = 1) {
  sliderInput(id, label, min = min, max = max, value = 0.5, step = 0.1)
}
```
We then call the function with `pmap()`
```{r}
sliders <- pmap(vars, mySliderInput)
ui <- fluidRow(sliders)
```


## Server Functions
When you have a long reactive, separating it into a function that does not use reactivity has two advantages:
* it's easier to debug and test the code when reactivity and computation are separate
* when looking at a reactive expression or output, there's no easy way to know exactly what values it depends on. A function will tell you exactly what the inputs are though

When writing functions from reactives, avoid taking reactives as input or returning outputs (unless necessary). Instead, pass values into arguments and assume the caller will turn the result into a reactive if needed.

Not every line in the reactive needs to go into a separate function. Certain parts of the reactive might be better suited to remaining in the reactive while other lines go in a function(s).

### Internal functions
Generally, you'll want to make the function completely independent of the server function so that you can put it in a separate file. However, if the function needs to use input, output or session, it may make sense for the function to live inside the server function.