---
title: "Chapter 19: Modules"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For code that spans both the UI and server, modules are required to decompose the app rather than functions.

At the simplest level, a module is a part of UI and server functions. The magic of modules is that the functions are constructed in a special way that creates a "namespace". Normally, when we write an app, the names (i.e. ids) are global: all parts of your server function can see all parts of the UI. Modules give you the ability to create controls that can only be seen from within the module. This is called a namespace.

Modules have two big advantages:
* namespacing, which makes it easier to understand how the app works because you can write, analyse and test individual components in isolation
* modules are functions so they reduce duplication. Anything that can be done with a function can be with a module


## Module Basics
Example app (note: it's too simple for there to be any real benefit to modularisation):
```{r}
ui <- fluidPage(
  selectInput("var", "Variable", names(mtcars)),
  numericInput("bins", "bins", 10, min = 1),
  plotOutput("hist")
)
server <- function(input, output, session) {
  data <- reactive(mtcars[[input$var]])
  output$hist <- renderPlot({
    hist(data(), breaks = input$bins, main = input$var)
  }, res = 96)
}
```

A module is very similar to an app, it has a UI and a server function. The two functions have a standard form; they both take an ID argument and use it to namespace the module.

### Module UI
Two steps:
* put the UI code inside a function that has an ID argument
* wrap each existing ID in a call to `NS()`, i.e. `*input(id = NS(id, "name"))`

Example:
```{r}
histogramUI <- function(id) {
  tagList(
    selectInput(NS(id, "var"), "Variable", choices = names(mtcars)),
    numericInput(NS(id, "bins"), "bins", value = 10, min = 1),
    plotOutput(NS(id, "hist"))
  )
}
```

In this example, the components are wrapped in `tagList()`, which is a special type of layout function that allows you to bundle together multiple components without actually implying how they will be laid out. When `histogramUI()` is called it can then be wrapped in a layout function like `column()`, `fluidRow()`, etc

### Module server
The module server is wrapped inside another function, which has an id argument. The module server takes the id parameter as well as a function that looks like the regular server function. It would look like this:
```{r}
histogramServer <- function(id) {
  moduleServer(id, function(input, output, session) {
    data <- reactive(mtcars[[input$var]])
    output$hist <- renderPlot({
      hist(data(), breaks = input$bins, main = input$var)
    }, res = 96)
  })
}
```

The names of the objects in the input list are specified in the module's UI function.

Having two levels of functions helps to distinguish the argument to your module from the arguments to the server function. `moduleServer()` takes care of namespacing automatically. It essentially does NS(id, "name") without you needing to specify it explicitly.

### Updated app
After writing module UI and module server functions, it's good practice to write a function that uses them to generate an app that can be used for experimentation and testing:
```{r}
histogramApp <- function() {
  ui <- fluidPage(
    histogramUI("hist1")
  )
  server <- function(input, output, session) {
    histogramServer("hist1")
  }
  shinyApp(ui, server)  
}
```

Note: the module UI and module server are connected by having the same ID.

### Namespacing
The key idea that makes modules work is that the name of each control (the id) is now determined by two pieces:
* the first piece comes from the module user, whoever calls `histogramServer()`
* the second piece comes from the module author, the developer who wrote `histogramServer()`
This means that the author does not need to worry about clashing with other UI components created by the user.

Namespacing turns modules into black boxes. From outside the module, you cannot see any of the inputs, outputs or reactives inside it. If you want to take input from reactives elsewhere in the app, you'll need to pass them to the module function explicitly.

### Naming conventions
It's a good idea to name the module and all its related components after what it is/does. In the example here:
* histogram.R holds all the code for the module
* histogramUI is the module UI
* histogramServer is the module server
* histogramApp() is the app that can be used for experimentation and testing

One of the added bonuses of modules is that you don't have to worry about component's IDs clashing across modules


## Inputs and Outputs
Sometimes a module with only an ID argument to the module UI and server is useful because it allows you to isolate complex code in its own file. However, a lot of the time, the module UI and server will need additional arguments. These will give you greater control over the module's appearance, allowing you to use the same module in more places. The module UI is just a regular function so there's relatively little that will be different in terms of how it is used.

The module server function can be more complex.

### Getting started: UI input + server output
Example app module that allows the user to select a dataset from the built-in data provided by the datasets package.

Module UI:
```{r}
datasetInput <- function(id, filter = NULL) { # filter can take is.data.frame or is.matrix
  names <- ls("package:datasets")
  if (!is.null(filter)) {
    data <- lapply(names, get, "package:datasets")
    names <- names[vapply(data, filter, logical(1))]
  }
  
  selectInput(NS(id, "dataset"), "Pick a dataset", choices = names)
}
```
In the module UI, there is one additional argument `filter`, which allows you to limit the options to built-in datasets that are either data frames or matrices.

Module server:
```{r}
datasetServer <- function(id) {
  moduleServer(id, function(input, output, session) {
    reactive(get(input$dataset, "package:datasets"))
  })
}
```
The module server uses `get()` to retrieve the dataset with its name. Unlike a regular server function, the module server function returns a value, the last expression processed. This value should always be a reactive.

To use a module server that returns a value, you just have to capture its return value with `<-`. So, if you wanted to display the dataset:
```{r}
datasetApp <- function(filter = NULL) {
  ui <- fluidPage(
    datasetInput("dataset", filter = filter),
    tableOutput("data")
  )
  server <- function(input, output, session) {
    data <- datasetServer("dataset")
    output$data <- renderTable(head(data()))
  }
  shinyApp(ui, server)
}
```

Note: in this example `datasetApp()` takes a filter argument and passes it onto the module UI, making it easy to experiment with that input.

### Case study: selecting a numeric variable
Building a control that allows the user to select variables of specified type from a reactive dataset.

As we want the dataset to be reactive, we can't fill in the choices when we start the app so the module UI is:
```{r}
selectVarInput <- function(id) {
  selectInput(NS(id, "var"), "Variable", choices = NULL) 
}
```

The server function will have two arguments: the data to select variables from and a filter used to select which variables to list.

To keep the module server simple, we can extract the key idea into a helper function:
```{r}
find_vars <- function(data, filter) {
  names(data)[vapply(data, filter, logical(1))]
}
```

The module server uses `observeEvent()` to update `selectInput()` choices when the data change and returns a reactive that provides the values of the selected variable
```{r}
selectVarServer <- function(id, data, filter = is.numeric) {
  moduleServer(id, function(input, output, session) {
    observeEvent(data(), {
      updateSelectInput(session, "var", choices = find_vars(data(), filter))
    })
    
    reactive(data()[[input$var]])
  })
}
```

Joining the module UI and server together:
```{r}
selectVarApp <- function(filter = is.numeric) {
  ui <- fluidPage(
    datasetInput("data", is.data.frame), # module from line 116
    selectVarInput("var"),
    verbatimTextOutput("out")
  )
  server <- function(input, output, session) {
    data <- datasetServer("data") # module from line 130
    var <- selectVarServer("var", data, filter = filter)
    output$out <- renderPrint(var())
  }
  
  shinyApp(ui, server)
}
```

### Server inputs
When designing a module server, you need to think about who is going to provide the value for each argument: the programmer calling the module or the user of the app? (Is the value fixed or is it something the user can change?). This determines whether or not an argument should be reactive.

`stopifnot()` can be used to help verify this. E.g. we can check that data is reactive and filter is not reactive in the server:
```{r}
selectVarServer <- function(id, data, filter = is.numeric) {
  stopifnot(is.reactive(data)) # will only run if is.reactive evaluates to true, i.e. if data is reactive
  stopifnot(!is.reactive(filter)) # will only run if filter is not reactive
  
  moduleServer(id, function(input, output, session) {
    observeEvent(data(), {
      updateSelectInput(session, "var", choices = find_vars(data(), filter))
    })
    
    reactive(data()[[input$var]])
  })
}
```
Alternatively, it could be more useful to use if statements and `stop()` with a hand crafted error message.

Checking that module inputs are reactive or not helps you avoid a common problem when you mix modules with other input controls. `input$var` is not a reactive so when you pass an input value into a module, you'll need to wrap it in a `reactive()`.

### Case study: histogram
The key challenge of creating modules is creating functions that are flexible enough to be used in multiple places, but simple enough to be easily understood.

The UI:
```{r}
histogramOutput <- function(id) {
  tagList(
    numericInput(NS(id, "bins"), "bins", 10, min = 1, step = 1),
    plotOutput(NS(id, "hist"))
  )
}
```

The server:
```{r}
histogramServer <- function(id, x, title = reactive("Histogram")) {
  stopifnot(is.reactive(x))
  stopifnot(is.reactive(title))
  
  moduleServer(id, function(input, output, session) { # note: we can have a module inside a module
    output$hist <- renderPlot({
      req(is.numeric(x()))
      main <- paste0(title(), " [", input$bins, "]")
      hist(x(), breaks = input$bins, main = main)
    }, res = 96)
  })
}
```
The server module has two additional inputs:
* x - a variable to plot
* title - the title for the histogram
Both of these inputs are reactive. title has a default value and as we want it to be reactive, we wrap the default value in `reactive()`

The app (making use of modules defined in the chapter):
```{r}
histogramApp <- function() {
  ui <- fluidPage(
    sidebarLayout(
      sidebarPanel(
        datasetInput("data", is.data.frame),
        selectVarInput("var"),
      ),
      mainPanel(
        histogramOutput("hist")    
      )
    )
  )
  
  server <- function(input, output, session) {
    data <- datasetServer("data")
    x <- selectVarServer("var", data)
    histogramServer("hist", x)
  }
  shinyApp(ui, server)
} 
```

### Multiple outputs
It would be a nice feature if the name of the variable selected was in the title of the histogram. Currently, `selectVarServer()` only returns the value of the variable but we can update it to return both the name and the value:
```{r}
selectVarServer <- function(id, data, filter = is.numeric) {
  stopifnot(is.reactive(data))
  stopifnot(!is.reactive(filter))
  
  moduleServer(id, function(input, output, session) {
    observeEvent(data(), {
      updateSelectInput(session, "var", choices = find_vars(data(), filter))
    })
    
    list(
      name = reactive(input$var),
      value = reactive(data()[[input$var]])
    )
  })
}
```

We then just update the app:
```{r}
histogramApp <- function() {
  ui <- fluidPage(
    sidebarLayout(
      sidebarPanel(
        datasetInput("data", is.data.frame),
        selectVarInput("var"),
      ),
      mainPanel(
        histogramOutput("hist")    
      )
    )
  )
  
  server <- function(input, output, session) {
    data <- datasetServer("data")
    x <- selectVarServer("var", data)
    histogramServer("hist", x$value, x$name)
  }
  shinyApp(ui, server)
} 
```

The challenge with this approach is remembering/knowing when you use the reactive (`x$value`) vs when you use its value (`x$value()`). The trick is remembering that when passing an argument to a module, you want the module to react to the value changing which means that you have to pass the reactive, not its current value.


## Case Studies
Module inputs: additional arguments to the module server can be reactives or constants. The choice is a design decision that you make based on who sets the arguments and when they change. You should always check that the arguments are of the expected type to avoid unhelpful error messages (e.g. `stopifnot()`).

Return values: unlike app servers, modules servers can return values. The return value of a module should always be a reactive, or a list of reactives (as demonstrated in Multiple outputs).

### Limited selection + other
An important use case of modules is giving complex UI elements a simpler UI. Here we create a small set of options displayed with radio buttons coupled with an "other" field. The inside of this module uses multiple input elements but from the outside it works as a single combined object.

```{r}
radioExtraUI <- function(id, label, choices, selected = NULL, placeholder = "Other") {
  other <- textInput(NS(id, "other"), label = NULL, placeholder = placeholder)
  
  names <- if (is.null(names(choices))) choices else names(choices)
  values <- unname(choices)
  
  radioButtons(NS(id, "primary"), 
    label = label,
    choiceValues = c(names, "other"),
    choiceNames = c(as.list(values), list(other)),
    selected = selected
  )
}
```
The UI has extra inputs/parameters in: label, choices, selected and placeholder. Label, choices and selected are passed onto `radioButtons()`. `textInput()` is used to create a placeholder that defaults to "Other". To combine the text box and radio button, we take advantage of the fact that `choiceNames` can be a list of HTML elements, including, another input widget. So, essentially, we artificially create an other option by creating a `textInput()` element that can be rendered as a `choiceName`. Note: if the user selects the "Other" radiobutton, `radioButtons()` will return "other" as the input value, not the value typed into `textInput()`. This is dealt with in the server.

```{r}
radioExtraServer <- function(id) {
  moduleServer(id, function(input, output, session) {
    observeEvent(input$other, ignoreInit = TRUE, { # ignoreInit tells observeEvent to ignore the handler expression
      updatepdateRadioButtons(session, "primary", selected = "other") # changes the name of the radio button
    })
    
    reactive({
      if (input$primary == "other") {
        input$other
      } else {
        input$primary
      }
    })
  })
}
```
When the user selects "Other", we use a reactive and an if statement to take the value they type into `textInput()` as the value of the server module.

Put together, it looks like this:
```{r}
radioExtraApp <- function(...) { # ... is there to pass more arguments to radioExtraUI() is needed
  ui <- fluidPage(
    radioExtraUI("extra", ...),
    textOutput("value")
  )
  server <- function(input, output, server) {
    extra <- radioExtraServer("extra")
    output$value <- renderText(paste0("Selected: ", extra()))
  }
  
  shinyApp(ui, server)
}
```


### Wizard
A wizard interface is a style of UI where you break a complex process down into a series of simple pages that the user works through one-by-one. The creation of this wizard will be automated so that you do not need to worry about how each page is connected.

The main part of the wizard UI are the buttons:
```{r}
nextPage <- function(id, i) {
  actionButton(NS(id, paste0("go_", i, "_", i + 1)), "next") # each input element needs to have a unique ID, the unique ID includes the current page's ID and the ID of the destination page (e.g. go_1_2)
}
prevPage <- function(id, i) {
  actionButton(NS(id, paste0("go_", i, "_", i - 1)), "prev")
}
```
Each page has two buttons: one to take the user to the next page and one to return them to the previous page.

The wizard page:
```{r}
wrapPage <- function(title, page, button_left = NULL, button_right = NULL) {
  tabPanel(
    title = title, 
    fluidRow(
      column(12, page)
    ), 
    fluidRow(
      column(6, button_left),
      column(6, button_right)
    )
  )
}
```
The wizard page has a title, its contents (page) and the two buttons.

The whole UI:
```{r}
wizardUI <- function(id, pages, doneButton = NULL) {
  stopifnot(is.list(pages))
  n <- length(pages)
  
  wrapped <- vector("list", n) # create a list for the number of pages desired
  for (i in seq_along(pages)) {
    # First page only has next; last page only prev + done
    lhs <- if (i > 1) prevPage(id, i)
    rhs <- if (i < n) nextPage(id, i) else doneButton
    wrapped[[i]] <- wrapPage(paste0("page_", i), pages[[i]], lhs, rhs) # create the contents of each page
  }
  
  # Create tabsetPanel
  # https://github.com/rstudio/shiny/issues/2927
  wrapped$id <- NS(id, "wizard")
  wrapped$type <- "hidden" # we want the tabs to be hidden so the user cannot skip pages
  do.call("tabsetPanel", wrapped) # is functionally the same as writing out `tabsetPanel(wrapped[[1]], wrapped[[2]], ..., id = NS(id, "wizard"))`
}
```
We loop over a list of supplied pages, create the buttons, then wrap up the supplied page into a `tabPanel()` and combine the panels together into `tabsetPanel()`.


With the server, we just need to make the buttons work so that users of the app can move from one page to another.
The server:
```{r}
wizardServer <- function(id, n) {
  moduleServer(id, function(input, output, session) {
    changePage <- function(from, to) { # define a function to change the page
      observeEvent(input[[paste0("go_", from, "_", to)]], {
        updateTabsetPanel(session, "wizard", selected = paste0("page_", to)) # changes the selected tab
      })  
    }
    ids <- seq_len(n)
    lapply(ids[-1], function(i) changePage(i, i - 1))
    lapply(ids[-n], function(i) changePage(i, i + 1))
  })
}
```
To carry out the function of the buttons, we need an `observeEvent()` for each button then call `updateTabsetPanel()`. But, as we don't know how many pages there will be in the app, we need to do some functional programming to set up (n-1)*2 observers, two observers for each page except the first and last, which only need one. `input[[]]` is used as a dynamic control to filter for the page we want. `lapply()` is used to loop over all the previous buttons, on all except the first page, and then again for all the next buttons, except for on the last page.


Putting it all together, we get:
```{r}
wizardApp <- function(...) {
  pages <- list(...)
  
  ui <- fluidPage(
    wizardUI("whiz", pages) # we don't create a done button here but it could be done with actionButton()
  )
  server <- function(input, output, session) {
    wizardServer("whiz", length(pages))
  }
  shinyApp(ui, server)
}
```
Note: we need to ensure that the `n` argument of `wizardServer()` is consistent with the `pages` argument of `wizardUI()` (i.e. n = the length of the pages list)


An example of putting the module into use:
```{r}
page1 <- tagList(
  textInput("name", "What's your name?")
)
page2 <- tagList(
  numericInput("age", "How old are you?", 20)
)
page3 <- tagList(
  "Is this data correct?",
  verbatimTextOutput("info")
)

ui <- fluidPage(
  wizardUI(
    id = "demographics", 
    pages = list(page1, page2, page3), 
    doneButton = actionButton("done", "Submit")
  )
)
server <- function(input, output, session) {
  wizardServer("demographics", 3) # n argument equals the length of pages

  observeEvent(input$done, showModal( # after the wizard server runs, we run something to tell the user that it has finished
    modalDialog("Thank you!", footer = NULL)
  ))
  
  output$info <- renderText(paste0( 
    "Age: ", input$age, "\n",
    "Name: ", input$name, "\n"
  ))
}
```