---
title: "Databases and Data APIs"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Databases and data APIs are just sources of data.


#### Accessing and using databases
Databases are defined by their query-able interface, usually through SQL. Any database connection starts with creating a connection object and you’ll use this object to send SQL queries (which can be done with {sqlalchemy} in Python or {dblyr} in R).
  
*	In Python, packages for individual databases directly implement the API so we just call the .connect() method
*	In R, the API is split into two parts. {DBI} implements the connection by working with database driver package, which is the first argument to DBI::dbConnect(). Note: not every database driver package in R is DBI-compliant
*	If there is not a database-specific package, you’ll need to use a generic system driver with a package to interface with the system driver
*	Data source name (DSN) holds the details of the connection, which makes connecting to a database straightforward as you don’t need to remember ports, usernames, passwords, etc
*	System drivers come in two main varieties: Java Database Connectivity (JDBC) and Open Database Connectivity (ODBC). Both have packages in R and Python for making connections



#### Providing credentials to data projects
Credentialling

*	Discretionary access control (DAC) is where you’re allowed to share data access permissions with others in your company
*	Service account is a non-human account that exists to hold permissions for a project. It can be used to limit the project’s permissions to exactly what it needs or to manage the project’s permissions independently of the humans involved
*	Using credentials of the person viewing the content and pass those along (the hardest option)
* This can be a complex process that requires work with/from IT


#### Connecting to APIs
Packages that wrap APIs work in a similar way to using a database – you create a connection object that stores the connection details. If you are calling an API directly, use the {requests} package in Python and {httr2} in R.

What is an API

*	In this context, http-based REST-ful APIs
*	HTTP operates on a request-response model – you send a request and the API sends back a response


API endpoints and paths

*	APIs have many endpoints, each of which lives on a path (looks like a directory)
*	Each request to an API is directed to a specific endpoint
*	The different endpoints can be sent different parameters and send back different things


HTTP verbs/request methods

*	Describes the type of operation you’re asking for
*	Each endpoint has one or more verbs that it knows how to use
*	GET – fetches information from the server
* POST – provides information to the server
*	PUT and PATCH – changes/updates
*	DELETE - deletes
* Request parameters and bodies
  * Each endpoint accepts specific arguments; some are optional, some are required
  * GET - arguments are specified via query parameters embedded in the URL after `?`
  * POST, PUT and PATCH - arguments are provided in a body, usually formatted as JSON. {httr2} and {requests} have functionality to convert standard R and Python data types to JSON equivalents. APIs often require arguments to be nested in specific ways
* Auth headers
  * most APIs require authentication
  * common forms of authentication: username and password, API keys and OAuth tokens
  * API keys and OAuth tokens ofter have scopes that describe what they have permission to do (e.g. GET)
* Request status codes
  * indicate what happened with your request to the server
  * 200 = success
  * 400s = problem with the request and the API couldn't understand what you wanted
  * 500s = request was fine but some error occurred in processing
* Response bodies
  * Bodies contain the contents of the response
  * Usually come in JSON format so {json} and {jsonlite} needed to decode them
* Common API patterns
  * pagination - paginated APIs return only a certain number of results to keep sizes modest
  * job APIs - when you have requested a long-running job, the API may return a job-id, which you can use to poll a job-status endpoint to check on progress and claim your result
  * multiple verbs - an endpoint accepting multiple verbs



#### Environment variables to secure data connections

When you take an app to production, authenticating your data source while keeping your secrets secure is crucial

* Never put your credentials in your code
* The easiest way to achieve this is with an environment variable - these are set before your code starts
* Getting environment variables
  * Python
    * read environment variables from the `os.environ` dictionary or by using `os.getenv("VAR_NAME")`
  * R
    * get environment variables with `Sys.getenv("VAR_NAME)`
  * note: convention is to make environment variable names all caps with underscores separating words
* Setting environment variables
  * most common way to set environment variables is to lead secrets from a text file
  * Python
    * we usually read an `.env` file in the session
    * {python-dotenv} is a good choice for doing this
  * R
    * automatically read the `.Renviron` file as environment variables and also sources the `.Rprofile` file where you can set environment variables using `Sys.setenv()`
  * {keyring}
    * a package in R and Python that allows you to use the system keyring to securely store environment variables and recall them at runtime
  * in production, setting environment variables is harder and is covered later


#### Data Connection Packages

* Contain all the shared connection details for an organisation so that users don't have to remember them
* Whether it's made in R or Python, the function in your package should return the database connection object for people to use
* Example if you were using a Postgres database from R:
```{r}
#' Return a database connection
#'
#' @param user username, character, defaults to value of DB_USER
#' @param pw password, character, defaults to value of DB_PW
#' @param ... other arguments passed to 
#' @param driver driver, defaults to RPostgres::Postgres
#'
#' @return DBI connection
#' @export
#'
#' @examples
#' my_db_con()
my_db_con <- function(
    user = Sys.getenv("DB_USER"), 
    pw = Sys.getenv("DB_PW"), 
    ..., 
    driver = RPostgres::Postgres()
) {
  DBI::dbConnect(
    driver = driver,
    dbname = 'my-db-name', 
    host = 'my-db.example.com', 
    port = 5432, 
    user = user,
    password = pw, 
    ...
  )
}
```

