---
title: 'Chapter 4: ER Injuries'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We explore a data set from the National Electronic Injury Surveillance System. 
```{r}
#---- load libraries and data ----
library(tidyverse)
library(vroom)

injuries <- vroom("neiss/injuries.tsv")
population <- vroom("neiss/population.tsv")
products <- vroom("neiss/products.tsv")
```

We pick a product and explore the patterns in the data
```{r}
# injuries that involve the toilet
toilet_injuries <- injuries |>
  filter(prod_code == 649)

# most injuries involving the toilet occur at home
toilet_injuries |>
  count(location, wt = weight, sort = T)

# the head is most often hurt in injuries involving the toilet
toilet_injuries |>
  count(body_part, wt = weight, sort = T)

# many different diagnoses
toilet_injuries |>
  count(diag, wt = weight, sort = T)

# injuries become more frequent in women than in men past the age of ~50
ggplot(toilet_injuries |>
         count(age, sex, wt = weight)) +
  geom_line(aes(age, n, colour = sex)) +
  ylab("Estimated number of injuries") +
  theme_classic()


# after standardising for population size, we see an increase in injuries for men and women past 50 but at around 80, injuries skyrocket for women
pop_adjusted_toilet_injuries <- toilet_injuries |>
  count(age, sex, wt = weight) |>
  left_join(population, by = c("age", "sex")) |>
  mutate(rate = n/population*10000)

ggplot(pop_adjusted_toilet_injuries) +
  geom_line(aes(age, rate, colour = sex), na.rm = T) +
  ylab("Estimated number of injuries") +
  theme_classic()

# we can sample the data to develop hypotheses
toilet_injuries |>
  slice_sample(n = 10) |>
  pull(narrative)
```

Having done that exploration, we want to explore the data for other products but there are a number of products. We could use a shiny app to do this easily.


## Prototype
Before developing a shiny app, it's a good idea to work out the basic mechanics first. 